Pigeons.jl: Distributed sampling from intractable distributions

### 2023-10-30 paper.tex
Remove inputs and place in main tex

\input{src/introduction.tex}
\input{src/examples.tex}
\input{src/PT.tex}
\input{src/SPI.tex}
\input{src/related.tex}
% \input{src/benchmarks.tex}
\input{src/conclusion.tex}

\section{Acknowledgements}
NS acknowledges the support of a Vanier Canada Graduate Scholarship. 
PT acknowledges the support of the Black Hole Initiative at Harvard University, 
which is funded by grants from the John Templeton Foundation and the 
Gordon and Betty Moore Foundation to Harvard University, 
as well as support by the National Science Foundation grants AST-1935980 
and AST-2034306 and the Gordon and Betty Moore Foundation (GBMF-10423).
SS acknowledges the support of EPSRC grant EP/R034710/1 CoSines.
ABC and TC acknowledge the support of an NSERC Discovery Grant.
We also acknowledge use of the ARC Sockeye computing platform from the
University of British Columbia, as well as cloud computing resources provided 
by Oracle.

>>>>

\section{Introduction}
In many scientific application domains, the ability to obtain samples from a 
distribution is crucial. 
For instance, sampling methods have been used to discover magnetic polarization 
in the black hole of galaxy M87 \cite{akiyama2021seven}
and to image the Sagittarius A* black hole \cite{akiyama2022first}.
They have also been used to 
model the evolution of single-cell cancer genomes \cite{salehi2021clonal}, 
infer plasma dynamics inside nuclear fusion reactors \cite{gota2021overview}, 
and to identify gerrymandering in Georgia's 2021 congressional districting plan 
\cite{zhao2022mathematically}.
Similarly, evaluating high-dimensional integrals or sums over complicated 
combinatorial spaces are related tasks that can also be solved with sampling 
via Markov chain Monte Carlo (MCMC) methods. 
However, such calculations can often be bottlenecks in the scientific process, with 
simulations that can last days or even weeks to finish.

\medskip
Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is currently compatible with Julia 1.8+.}
enables users to sample efficiently from high-dimensional and complex distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple API allows users to perform such computations 
single-threaded, multi-threaded, and/or distributed over thousands of MPI-communicating 
machines. Further, it comes with guarantees on strong parallelism invariance wherein 
the output for a given seed is \emph{identical} irrespective 
of the number of threads or processes. 
Such a level of reproducibility is rare in distributed software but of 
great use for the purposes of debugging in the context of sampling algorithms, 
which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

\subsection{Problem formulation}
We describe the class of problems that can be approached using Pigeons.jl.
Let $\pi(x)$ denote a probability density
\footnote{This density may also be a probability mass function 
(discrete variables). We also allow for a combination of discrete/continuous variables.} 
called the \emph{target}. 
In many problems, e.g. in Bayesian statistics, the density $\pi$ is typically 
known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.

\medskip 
The output of Pigeons.jl can be used for two main tasks:
\begin{enumerate}
    \item Approximating integrals of the form $\int f(x) \pi(x) \, \dee x$.  
    For example, the choice $f(x) = x$ computes the mean and 
    $f(x) = I[x \in A]$ computes the probability of $A$ under $\pi$,
    where $I[\cdot]$ denotes the indicator function.

    \item Approximating the value of the normalization constant $Z$. For 
    example, in Bayesian statistics, this corresponds to the 
    marginal likelihood, which can be used for model selection. 
\end{enumerate}
Its implementation shines compared to traditional sampling packages in the 
following scenarios:
\begin{itemize}
    \item When the target density $\pi$ is challenging due to a complex structure 
    (e.g., high-dimensional, multi-modal, etc.).

    \item When the user needs not only $\int f(x) \pi(x) \, \dee x$ but also
    the normalization constant $Z$. 
    Many existing tools focus on the former and struggle or fail to do the latter.

    \item When the target distribution $\pi$ is defined over a non-standard space, 
    e.g. a combinatorial object such as a phylogenetic tree.  
\end{itemize}

\subsection{What is of interest to the general Julia developer?}
Ensuring code correctness at the intersection of randomized, parallel, and 
distributed algorithms is a challenge. To address this challenge, we designed Pigeons.jl 
based on a principle that we refer to as \textit{strong parallelism invariance}
(SPI).
Namely, the output of Pigeons.jl is completely invariant to the number of machines 
and/or threads.
Without explicitly keeping SPI in mind during software construction, 
the (random) output of the algorithm is only guaranteed to have the same distribution.
This is a much weaker guarantee that, in particular, makes debugging difficult.
However, with our notion of SPI we make debugging and 
software validation considerably easier. This is because the developer can 
first focus on the fully serial randomized algorithm, and then use it as an 
easy-to-compare gold-standard reference for parallel/distributed implementations. 
This strategy is used extensively in Pigeons.jl to ensure correctness. In contrast, 
testing equality in distribution, while possible (e.g., see \cite{geweke2004getting}), 
incurs additional false negatives due to statistical error.

\medskip 
The general Julia developer will be interested in: 
\begin{itemize}
    \item The main causes of a violation of strong parallelism invariance that we have identified 
    (\cref{sec:PI_causes})---some of which are specific to Julia---and how we address 
    them in Pigeons.jl.

    \item The SplittableRandoms.jl\footnote{\url{https://github.com/Julia-Tempering/SplittableRandoms.jl}}
    package that was developed by our team 
    to achieve strong parallelism invariance in Pigeons.jl (\cref{sec:splittable_randoms}).
\end{itemize}

\section{Examples}
In this section we present a set of minimal examples 
that demonstrate how to use Pigeons.jl for sampling. 
For further reading, we also direct readers to our growing list of examples, which can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl/tree/main/examples}.
We begin by installing the latest official release of Pigeons.jl: 
\begin{lstlisting}[language = Julia]
using Pkg; Pkg.add("Pigeons")
\end{lstlisting}

\subsection{Targets}
To use Pigeons.jl, we must specify a target distribution, given by $\gamma$ 
in \cref{eq:normalizing_constant}.
Numerous possible types of target distributions are supported, including 
custom probability densities (specified up to a normalizing constant) written in Julia.
We also allow to interface with models written in common probabilistic programming 
languages, including:
\begin{itemize}
    \item Turing.jl \cite{ge2018turing} models (\texttt{TuringLogPotential})
    \item Stan \cite{carpenter2017stan} models (\texttt{StanLogPotential})
    \item Comrade.jl\footnote{\url{https://github.com/ptiede/Comrade.jl}} 
      models for black hole imaging (\texttt{ComradeLogPotential})
    \item Non-Julian models with foreign-language Markov chain Monte Carlo (MCMC) code 
    (e.g. Blang \cite{bouchard2022blang} code for phylogenetic inference over combinatorial spaces) 
\end{itemize}
Additional targets are currently being accommodated and will be 
introduced to Pigeons.jl in the near future.

\medskip 
In what follows, we demonstrate how to use Pigeons with a Julia Turing model 
applied to a non-identifiable ``coinflip'' data set, modified from the example
at \url{https://turing.ml/v0.22/docs/using-turing/quick-start}. 
The Bayesian model can be formulated as 
\[
  \label{eq:coinflip}
  p_1, p_2 &\stackrel{iid}{\sim} U(0, 1) \\    
  Y \mid p_1, p_2 &\sim \text{Binomial}(n, p_1 p_2).
\]
The random variable $Y$ is the number of heads observed on $n$ coin flips
where the probability of heads is $p_1 p_2$.  
This model is non-identifiable, meaning that it is not possible to distinguish 
the effects of the two different parameters $p_1$ and $p_2$. As a consequence, 
the target distribution exhibits a complicated structure, as displayed in 
\cref{fig:coinflip_posterior}.
The density of interest corresponding to this model is 
\[
  \pi(p_1, p_2) &= \gamma(p_1, p_2)/Z,
\] 
where 
\[ 
  \gamma(p_1, p_2) &= 
    {n \choose y} (p_1 p_2)^y (1-p_1 p_2)^{n-y} I[p_1, p_2 \in [0,1]] 
    \label{eq:coinflip_density} \\
  Z &= \int_0^1 \int_0^1 \gamma(p_1, p_2) \, \dee p_1 \, \dee p_2. \label{eq:coinflip_normalization}
\]
The distribution $\pi$ is also known as the \emph{posterior distribution} in 
Bayesian statistics.

\begin{figure}[t]
    \centering 
    \includegraphics[width=0.5\textwidth]{img/coinflip_posterior.pdf}
    \caption{Posterior distribution for the model given by \cref{eq:coinflip} 
    with $n=100,000$ coin flips and $y=50,000$ observed heads, 
    estimated using $2^{17}$ samples from Pigeons.jl. 
    We present the pairwise plot for $p_1$ and $p_2$, as well as the estimated 
    densities of the marginal of the posterior for each of the two parameters.
    Note that because the model is non-identifiable, as we collect more data the 
    posterior distribution concentrates around the curve $p_1 p_2 = 0.5$, instead 
    of a single point, assuming that the true probability of observing heads 
    is 0.5.}
    \label{fig:coinflip_posterior}
\end{figure}

\medskip 
Suppose that we perform $n=100,000$ coin tosses and observe 
$y=50,000$ heads.
We would like to obtain samples from our posterior, $\pi$, having collected this data.
We begin by installing Turing
\begin{lstlisting}[language = Julia]
Pkg.add("Turing")
\end{lstlisting}
and then defining our Turing model and storing it in the variable \texttt{model}:
\begin{lstlisting}[language = Julia]
using Turing
@model function coinflip(n, y)
    p1 ~ Uniform(0.0, 1.0)
    p2 ~ Uniform(0.0, 1.0)
    y ~ Binomial(n, p1 * p2)
    return y
end
model = coinflip(100000, 50000)
\end{lstlisting}

From here, it is straightforward to sample from the density given by 
\cref{eq:coinflip_density} up to a normalizing constant.
We use non-reversible parallel tempering \cite{syed2021nrpt} (PT), Pigeons.jl's 
state-of-the-art sampling algorithm, to sample from the target distribution.
PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
to select these parameters effectively, which Pigeons.jl implements under the hood.
We also specify that we would like to store 
the obtained samples in memory to be able to produce trace-plots, as well 
as some basic online summary statistics of the target distribution and useful 
diagnostic output by specifying 
\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.allocation\_extrema]}.
It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
will be selected for the user.
The code below runs Pigeons.jl on one machine with one thread.
We use the default values for most settings, however 
we explain later how one can obtain improved performance by setting 
arguments more carefully (see \cref{sec:additional_options}).
\begin{lstlisting}[language = Julia]
using Pigeons
pt = pigeons(
    target = TuringLogPotential(model), 
    record = [
        traces, online, round_trip, 
        Pigeons.timing_extrema, 
        Pigeons.allocation_extrema])
\end{lstlisting}
Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
we pass the model as an argument to \texttt{TuringLogPotential()}.
Once we have stored the PT output in the variable \texttt{pt} we can 
access the results, as described in the following section. 
The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
and explained in the next section. 
For purposes of comparison, we also run a traditional 
(single-chain Markov chain Monte Carlo) method.

\begin{figure*}[t]
    \centering
    \begin{lstlisting}
          --------------------------------------------------------------------------------------
            #scans    restarts      Λ        time(s)    allc(B)  log(Z₁/Z₀)   min(α)     mean(α)
          ---------- ---------- ---------- ---------- ---------- ---------- ---------- --------- 
                  2          0       1.04      0.383   3.48e+07  -4.24e+03          0      0.885
                  4          0       4.06    0.00287   1.79e+06      -16.3   4.63e-06      0.549
                  8          0       3.49    0.00622   3.55e+06      -12.1      0.215      0.612
                 16          0       2.68     0.0161   7.46e+06      -10.2      0.518      0.703
                 32          0       4.29     0.0353   1.37e+07      -11.8      0.222      0.524
                 64          3       3.17     0.0699   2.86e+07      -11.5      0.529      0.648
                128          8       3.56      0.139   5.53e+07      -11.5      0.523      0.605
                256         12       3.38      0.241    1.1e+08      -11.6      0.526      0.625
                512         37       3.48      0.473   2.22e+08        -12      0.527      0.614
           1.02e+03         77       3.55      0.895   4.46e+08      -11.8      0.571      0.605
          --------------------------------------------------------------------------------------
    \end{lstlisting}
    \caption{Standard output provided by Pigeons.jl. 
    Rows indicate tuning rounds of the PT algorithm with an exponentially increasing number 
    of PT iterations (\texttt{\#scans}). Columns indicate various useful diagnostics, 
    such as the number of allocations per round, time (in seconds), and estimates 
    of the log of the normalization constants. 
    The output is described in greater detail in \cref{sec:PT_diagnostics}
    and has been modified to exclude columns that are not described in the paper.}
    \label{fig:standard_out}
\end{figure*}

\subsubsection{Other targets}
As mentioned previously, it is also possible to specify targets with custom 
probability densities, as well as Stan and Turing models. 
Additionally, suppose we have some code implementing vanilla MCMC, written 
in an arbitrary ``foreign'' language such as C++, Python, R, Java, etc. 
Surprisingly, it is very simple to bridge such code with Pigeons.jl. 
The only requirement on the foreign language is that it supports 
reading the standard input and writing to the standard output, 
as described in our online documentation.

\subsection{Outputs}
Pigeons.jl provides many useful types of output, such as: plots of samples from 
the distribution, estimates of normalization constants, summary statistics of 
the target distribution, and various other diagnostics. 
We describe several examples of possible output below.

\subsubsection{Standard output}
An example of the standard output provided by Pigeons.jl is displayed in \cref{fig:standard_out}. 
Each row of the table in the output indicates a new tuning round in parallel tempering, 
with the \texttt{\#scans} column indicating the number of scans/samples in 
that tuning round. 
During these tuning rounds, Pigeons.jl searches for optimal values of certain PT 
tuning parameters. Other outputs include: 
\begin{itemize}
    \item \texttt{restarts}: a higher number is better. 
    Informally, PT performs well at sampling from high-dimensional and/or multi-modal 
    distributions by pushing samples from an easy-to-sample distribution (the reference) 
    to the more difficult target distribution. 
    A tempered restart happens when a sample from the reference successfully 
    percolates to the target. (See the subsequent sections for a more detailed description 
    of parallel tempering.) 
    When the reference supports i.i.d.~sampling, tempered restarts 
    can enable large jumps in the state space.

    \item \texttt{$\Lambda$}: the global communication barrier, as described in \cite{syed2021nrpt}, 
    which measures the inherent difficulty of the sampling problem.
    A rule of thumb to configure the number of PT chains is also 
    given by \cite{syed2021nrpt}, where they suggest that stable performance 
    should be achieved when the number of chains is set to roughly 2$\Lambda$.
    See \cref{sec:PT_diagnostics} for more information.

    \item \texttt{time} and \texttt{allc}: the time (in seconds) and number of 
    allocations (in bytes) used in each round.

    \item \texttt{$\log(Z_1/Z_0)$}: an estimate of the 
    logarithm of the ratio of normalization constants between the target and the reference. 
    In many cases, $Z_0 = 1$.

    \item \texttt{min($\alpha$)} and \texttt{mean($\alpha$)}: minimum and average swap 
    acceptance rates during the communication phase across the PT chains.
    See \cref{sec:PT} for a description of PT communication.
\end{itemize}

\subsubsection{Plots}
It is straightforward to obtain plots of samples from the target distribution, 
such as trace-plots, pairwise plots, and density plots of the marginals.

\medskip 
To obtain posterior densities and trace-plots, we first 
make sure that we have the third-party 
MCMCChains.jl\footnote{\url{https://github.com/TuringLang/MCMCChains.jl}}, 
StatsPlots.jl\footnote{\url{https://github.com/JuliaPlots/StatsPlots.jl}}, and 
PlotlyJS.jl\footnote{\url{https://github.com/JuliaPlots/PlotlyJS.jl}} 
packages installed via
\begin{lstlisting}[language=Julia]
Pkg.add("MCMCChains", "StatsPlots", "PlotlyJS")
\end{lstlisting}

With the \texttt{pt} output object from before for our non-identifiable coinflip model, 
we can run the following:
\begin{lstlisting}[language=Julia]
using MCMCChains, StatsPlots, PlotlyJS
plotlyjs()
samples = Chains(
    sample_array(pt), variable_names(pt))
my_plot = StatsPlots.plot(samples)
display(my_plot)
\end{lstlisting}
The output of the above code chunk is an interactive plot that can be zoomed in or out
and exported as an HTML webpage. 
A modified static version of the output for the first parameter, $p_1$, 
is displayed in the top panel of \cref{fig:trace_density_plots},
along with a comparison to the output from a single-chain algorithm in the bottom 
panel of the same figure.

\begin{figure*}[t]
    \centering
    \begin{subfigure}{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/trace_density_plots_pt.pdf}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/trace_density_plots_single_chain.pdf}
    \end{subfigure}
    \caption{
        Trace-plots for the first parameter, $p_1$, in the 
        non-identifiable coinflip Turing model.
        \textbf{Left:} Samples from Pigeons.jl using PT with 10 chains. Note that the 
        trace-plot indicates fast mixing/exploration across the state space.
        \textbf{Right:} Single-chain Markov chain Monte Carlo. Note that the 
        trace-plot explores the state space much more slowly when we do not use PT.
    }
    \label{fig:trace_density_plots}
\end{figure*}

\medskip 
To obtain pair plots, we add the 
PairPlots.jl\footnote{\url{https://github.com/sefffal/PairPlots.jl}}
and CairoMakie.jl\footnote{\url{https://github.com/JuliaPlots/CairoMakie.jl}}
packages:
\begin{lstlisting}[language=Julia]
Pkg.add("PairPlots", "CairoMakie")
\end{lstlisting}
and then run
\begin{lstlisting}[language=Julia]
using PairPlots, CairoMakie
my_plot = PairPlots.pairplot(samples) 
display(my_plot)
\end{lstlisting}
The output of the above code chunk with an increased number of samples 
is displayed in \cref{fig:coinflip_posterior}.
\footnote{The code chunk above only works with Julia 1.9.}

\subsubsection{Estimate of normalization constant}
The (typically unknown) constant $Z$ in \cref{eq:normalizing_constant} is 
referred to as the \emph{normalization constant}. 
In many applications, it is useful to approximate this constant. 
For example, in Bayesian statistics, this corresponds to the 
marginal likelihood and can be used for model selection.

\medskip 
As a side-product of PT, we automatically obtain an approximation to the natural 
logarithm of the normalization constant. This is done automatically using the 
stepping stone estimator \cite{xie2011improving}.
The estimate can be accessed using
\begin{lstlisting}[language=Julia]
stepping_stone(pt)
\end{lstlisting}
In the case of the normalization constant given by \cref{eq:coinflip_normalization} 
for $n=100,000$ and $y=50,000$, we can exactly obtain its value as $\log(Z) \approx -11.8794$ 
using a computer algebra system. Note that this is very close to the output provided in 
\cref{fig:standard_out}.

\subsubsection{Online statistics}
\label{sec:online_stats}
Having specified the use of the \texttt{online} recorder in our call to \texttt{pigeons()}, 
we can output some basic summary statistics of the marginals of our target distribution. 
For instance, it is straightforward to estimate the mean and variance of each 
of the marginals of the target with 
\begin{lstlisting}[language=Julia]
using Statistics
mean(pt); var(pt)
\end{lstlisting}
Other constant-memory statistic accumulators are made available in the 
OnlineStats.jl \cite{day2020onlinestats} package. 
To add additional constant-memory statistic accumulators, we 
can register them via \texttt{Pigeons.register\_online\_type()}, as described in our 
online documentation. For instance, we can also compute constant-memory estimates 
of extrema of our distribution.

\subsubsection{Off-memory processing}
When either the dimensionality of the model or the number of samples is large,
the obtained samples may not fit in memory. 
In some cases it may be necessary to store samples to disk if our statistics of 
interest cannot be calculated online and with constant-memory
(see \cref{sec:online_stats}).
We show here how to save samples to disk when Pigeons.jl is run on a single 
machine. A similar interface can be used over MPI.

\medskip 
First, we make sure that we set \texttt{checkpoint = true}, which saves a 
snapshot at the end of each round in the directory \texttt{results/all/<unique directory>}
and is symlinked to \texttt{results/latest}.
Second, we make sure that we use the \texttt{disk} recorder 
by setting \texttt{record = [disk]}, along with possibly any other desired recorders.
Accessing the samples from disk can then be achieved in a simple way using the Pigeons.jl 
function \texttt{process\_sample()}.

\subsubsection{PT diagnostics}
\label{sec:PT_diagnostics}
We describe how to produce some key parallel tempering diagnostics from 
\cite{syed2021nrpt}.

\medskip 
The global communication barrier, denoted $\Lambda$ in Pigeons.jl output, can be 
used to approximately inform the appropriate number of chains. 
Based on \cite{syed2021nrpt}, stable PT performance should be achieved when  
the number of chains is set to roughly $2\Lambda$. This can be achieved by 
modifying the \texttt{n\_chains} argument in the call to \texttt{pigeons()}. 
The global communication barrier is shown at each round and can also be accessed 
with 
\begin{lstlisting}[language=Julia]
Pigeons.global_barrier(pt)
\end{lstlisting}

The number of restarts per round can be accessed with 
\begin{lstlisting}[language=Julia]
n_tempered_restarts(pt)
\end{lstlisting}
These quantities are also displayed in \cref{fig:standard_out}. 
Many other useful PT diagnostic statistics and plots can be obtained, as described 
in our full documentation.

\subsection{Parallel and distributed PT}
One of the main benefits of Pigeons.jl is that it allows users to easily parallelize 
and/or distribute their PT sampling efforts. We explain how to run MPI locally on 
one machine and also how to use MPI when a cluster is available.

\subsubsection{Running MPI locally}
To run MPI locally on one machine using four MPI processes and one thread per process,
use
\begin{lstlisting}[language = Julia]
pigeons(
    target = TuringLogPotential(model), 
    on = ChildProcess(
            n_local_mpi_processes = 4,
            n_threads = 1))
\end{lstlisting}

\subsubsection{Running MPI on a cluster}
Often, MPI is available via a cluster scheduling system. To run MPI over 
several machines:
\begin{enumerate}
    \item In the cluster login node, follow the Pigeons.jl installation instructions
    in our online documentation. 
    \item Start Julia in the login node, and perform a one-time setup by 
    calling \texttt{Pigeons.setup\_mpi()}.
    \item In the Julia REPL running in the login node, run:
\end{enumerate}
\begin{lstlisting}[language = Julia]
pigeons(
    target = TuringLogPotential(model), 
    n_chains = 1_000,
    on = MPI(n_mpi_processes = 1_000, 
             n_threads = 1))
\end{lstlisting}
The code above will start a distributed PT algorithm with 1,000 chains on 1,000 
MPI processes each using one thread.
Note that for the above code chunks involving \texttt{ChildProcess()} and 
\texttt{MPI()} to work, it may be necessary to specify dependencies in their 
function calls.

\subsection{Additional options}
\label{sec:additional_options}
In the preceding example we only specified the target distribution and let 
Pigeons.jl decide on default values for most other settings of the inference engine. 
There are various settings we can change, including: 
the random seed (\texttt{seed}), the number of PT chains (\texttt{n\_chains}), 
the number of PT tuning rounds/samples (\texttt{n\_rounds}), 
and a variational reference distribution family (\texttt{variational}), among other settings.
For instance, we can run 
\begin{lstlisting}[language = Julia]
pigeons(
    target = TuringLogPotential(model),
    n_rounds = 10, 
    n_chains = 10,
    variational = GaussianReference(),
    seed = 2
)
\end{lstlisting}
which runs PT with the same Turing model target as before and explicitly states 
that we should use 10 PT tuning rounds with 10 chains (described below). 
In the above code chunk we also specify that we would like to use a 
Gaussian variational reference distribution.
That is, the reference distribution is chosen from a multivariate Gaussian family 
that lies as close as possible to the target distribution in order to improve 
the efficiency of PT. We refer readers to \cite{surjanovic2022vpt} for more details.
When only continuous parameters are of interest, we encourage users to consider 
using \texttt{variational = GaussianReference()} and setting \texttt{n\_chains\_variational = 10}, 
for example, as the number of restarts may substantially increase with these settings.

\section{Parallel tempering}
\label{sec:PT}
Pigeons.jl provides an implementation of distributed parallel tempering (PT) described 
in \cite{syed2021nrpt}, which we outline in Algorithm~\ref{alg:distributed_PT}.
This section gives both a brief overview of PT and some details of 
its distributed implementation.

\medskip 
In this section we assume a basic understanding of 
Markov chain Monte Carlo (MCMC) methods.
For readers unfamiliar with MCMC, it is important to know that 
it is a method to obtain approximate samples from a distribution. 
More specifically, it is a class of algorithms for simulating a Markov chain of 
states that look like draws from the target distribution $\pi$.
However, with traditional 
MCMC methods, the samples may be heavily correlated instead of independent 
and may fail to sufficiently explore the 
full space of the distribution (see \cref{fig:bimodal}); 
PT is a method that aims to address these issues.
For a more in-depth review of PT, we refer readers to \cite{surjanovic2022vpt}.

\subsection{Overview of PT}
Suppose that we would like to estimate integrals involving $\pi$, such as 
$\int f(x) \pi(x) \, \dee x$. 
These integrals may be multivariate and even include combinations of 
continuous and discrete variables (where sums replace integrals in the discrete 
case).
One method is to obtain samples from $\pi$ to approximate such integrals.
Often, the distribution 
$\pi$ can be challenging for traditional MCMC methods---such 
as Metropolis-Hastings, slice sampling, and Hamiltonian Monte Carlo---
because of its structure.
For example, in a bimodal example such as the one illustrated in 
Figure~\ref{fig:bimodal},
traditional methods might remain in one of the two modes for an extremely 
long period of time.

\begin{figure}[t]
    \centering
    \includegraphics[width=0.3\textwidth]{img/bimodal.pdf}
    \caption{A simple bimodal distribution from which traditional 
    MCMC methods may struggle to obtain samples. Blue lines display output 
    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
    The sampler in this figure is visibly trapped in one of the two modes.}
    \label{fig:bimodal}
\end{figure}

To resolve this issue, PT constructs a sequence of $N$ distributions,  
$\pi_1, \pi_2, \ldots, \pi_N$, where $\pi_N$ is usually equal to $\pi$.
The distributions are chosen so that it is easy to obtain samples from $\pi_1$
with the sampling difficulty increasing as one approches $\pi_N$. An example 
of such a sequence of distributions, referred to as an \textit{annealing path},
is shown in Figure~\ref{fig:path}.

\begin{figure}[t]
    \centering
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_1.pdf}
      \caption*{$\pi_1$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_2.pdf}
      \caption*{$\pi_2$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_3.pdf}
      \caption*{$\pi_3$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_4.pdf}
      \caption*{$\pi_4$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_5.pdf}
      \caption*{$\pi_5$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_6.pdf}
      \caption*{$\pi_6$}
    \end{subfigure}
    \caption{Heatmaps of six distributions lying on an annealing path 
    from a unimodal reference distribution  
    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
    $\pi_N$, which is in this case the bimodal distribution from 
    Figure~\ref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
    are not directly comparable because the densities of intermediate distributions 
    are not normalized.)}
    \label{fig:path}
\end{figure}

We now turn to explain how this path of distributions can be used to enhance 
sampling from the target distribution, $\pi$. 
PT operates by first obtaining samples from each distribution on the path in parallel 
(referred to as an \textit{exploration phase}). 
Then, samples between adjacent distributions are swapped (referred to as 
a \textit{communication phase}). The communication phase in PT is crucial: it allows 
for the discovery of new regions of the space of the target distribution such as 
the top-right mode of the distribution presented in Figure~\ref{fig:bimodal}.

\subsection{Local exploration and communication}
A struct that is useful for the implementation of PT is the \texttt{Replica}, 
which we will often refer to in the remainder of this section. 
A single \texttt{Replica} struct stores a \texttt{state} variable and a \texttt{chain} 
integer, among other entries. 
At the beginning of PT, $N$ \texttt{Replicas} are created, 
one for each distribution on the annealing path, and 
the chain entries in the $N$ replicas are initialized at $1,2,\ldots,N$, respectively. 
For a given \texttt{Replica}, if the \texttt{chain} number is $i$ and the \texttt{state}
is $x$, then this means that the sample corresponding to the $i$-th distribution in 
the sequence $\pi_1, \ldots, \pi_N$ is currently at location $x$.

\medskip 
In the local exploration phase,
each \texttt{Replica}'s state is modified using an MCMC move targeting $\pi_i$,
where $i$ is given by \texttt{Replica.chain}.
The MCMC move can either modify \texttt{Replica.state} in-place, or modify the 
\texttt{Replica}'s \texttt{state} field. 
This operation is indicated by the \texttt{local\_exploration} function in 
\cref{alg:distributed_PT}.

\medskip 
In the communication phase, PT proposes swaps between pairs of replicas. 
In principle, there are two equivalent ways to do a swap.
In the first implementation, the \texttt{Replica}s 
could exchange their \texttt{state} fields.
Alternatively, they could exchange their \texttt{chain} fields.
Because we provide distributed implementations, we use the latter as it ensures that 
the amount of data that needs to be exchanged between two machines during a swap 
can be made very small (two floating point numbers), resuling in 
an exchange of $O(N)$ messages of size $O(1)$. 
Note that this cost does not vary with the dimensionality of the state space, 
in contrast to the first implementation that would transmit 
$O(N)$ messages of size $O(d)$, where $d$ is the dimension of the state space,
incurring a potentially very high communication cost for large values of $d$.

\subsection{Distributed implementation}
A distributed implementation of PT is presented in \cref{alg:distributed_PT}
and we describe the details of the implementation below.
We present the algorithm from the perspective that the number of machines available 
is equal to the number of \texttt{Replica}s and distributions, $N$. 
However, Pigeons.jl also allows for the more general case where the number of 
machines is not necessarily equal to $N$.

\subsubsection{The \texttt{PermutedDistributedArray}}
\label{sec:permuted_dist_array}
Recall that our theoretical $O(1)$ message size for communication between two machines 
is achieved by exchanging \texttt{chain} indices between \texttt{Replica}s 
instead of their \texttt{state}s.
One difficulty can be encountered in a distributed implementation, which we 
illustrate with the following simple example. 
Suppose we have $N=4$ distributions, chains, replicas, and machines, and that 
machine 1 is exploring chain 2 while machine 4 is exploring chain 3.
At one point, the \texttt{Replica} at \texttt{chain} 2 (machine 1) might need to exchange
\texttt{chain} indices with the \texttt{Replica} that has \texttt{chain} index 3 (machine 4). 
However, a priori it is not clear how machine 1 should know that it should communicate 
with machine 4 because it has no knowledge about the chain indices on the other machines.

\medskip 
To resolve this issue, we introduce a special data structure called a 
\texttt{PermutedDistributedArray}. In the case where the number of replicas is 
equal to the number of available machines, the construction is quite simple
and effectively results in each machine storing one additional integer. 
Considering the same example above, the solution to the problem is to have
machine 1 (\texttt{chain} 2 wanting to communicate with \texttt{chain} 3) 
communicate with \textit{machine} 3. 
Machine 3 stores in its \texttt{dist\_array} variable of type 
\texttt{PermutedDistributedArray} the value 4, which is the machine number that 
stores \texttt{chain} 3. By updating these \texttt{PermutedDistributedArray} 
variables at each communication step, we can ensure that each machine $j$ is aware 
of which machine number currently stores \texttt{chain} $j$. 
Therefore, the machines act as keys in a dictionary for the communication 
permutations.
An illustration of communication between four machines is provided in 
\cref{fig:index_process}.
We note that with a \texttt{PermutedDistributedArray} we make special assumptions 
on how we access and write to the array elements. 
Several MPI processes cooperate, with each machine storing 
data for a slice of this distributed array, and at each time 
step an index of the array is manipulated by exactly one machine.
% With these assumptions we achieve read/write costs that are 
% near-constant in the number of machines participating.
